---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

First things first: Loading needed libraries
```{r load_libraries, warning=FALSE, message=FALSE}
library("dplyr")
library("ggplot2")
library("readr")
library("stringr")
library("httr")
library("jsonlite")
library("tidyverse")
library("coloc")
library("biomaRt")
library("wiggleplotr")
library("GenomicRanges")
library("biomaRt")
```

Make an output directory for all of the figures
```{r, warning =FALSE}
dir.create("eQTL_API_usecase_figures")
```

We defined a local function to easily fetch data from both GWAS and eQTL APIs
```{r define_function_to_fetch_paginated_API_response, warning=FALSE}
fetch_from_eqtl_cat_API <- function(link, is_gwas = FALSE){
  nullToNA <- function(x) {
    x[sapply(x, is.null)] <- NA
    return(x)
  }
  if (is_gwas) {
    cols_to_nest <- c("variant_id", "chromosome", "base_pair_location", "trait", 
                      "p_value", "ci_lower", "ci_upper", "beta", 
                      "effect_allele", "other_allele", "effect_allele_frequency", 
                      "odds_ratio", "study_accession", "code")
  } else {
    cols_to_nest <- c("study_id", "qtl_group", "rsid", 
                       "chromosome", "position", "pvalue", "condition_label", 
                       "tissue_label", "molecular_trait_id", "gene_id", "ac",  
                       "ref", "beta",  "variant", "an", "median_tpm",  "condition", 
                       "r2", "alt", "type", "maf",  "tissue")
  }
  is_paginated <- !str_detect(link,"paginate=False")
  # message("isPagined:", is_paginated)
  page = 1
  merged_summaries <- data.frame()
  while(!is.null(link)){
    # print(paste0("Fetching page #",page))
    api_raw_data <- fromJSON(link, simplifyDataFrame = TRUE, flatten = TRUE)
    link <- api_raw_data$`_links`$`next`$href
    if (is_empty(api_raw_data$`_embedded`$associations)) {
      return(merged_summaries)
    }
    eqtl_raw_list_data <- do.call(rbind, lapply(api_raw_data$`_embedded`$associations, rbind))
    eqtl_data <- nullToNA(eqtl_raw_list_data) %>% as.matrix() %>% as_tibble()
    if (is_paginated) { eqtl_data <- dplyr::select(eqtl_data, -c("_links")) }
    eqtl_data <- tidyr::unnest(eqtl_data, cols = cols_to_nest)
    if (!is.null(link)) {
      page <- page + 1
    }
    merged_summaries <- merged_summaries %>% rbind(eqtl_data)
  }
  return(merged_summaries[cols_to_nest])
}
```

Method to get significant associations with lead GWAS variant ID.
```{r}
get_significant_assocs_of_var <- function(study_ids, variant_id, p_upper = 0.0001, quant_method = "ge"){
  rnaseq_studies_sign_assocs <- data.frame()
  for(study_name in study_ids){
    print(study_name)
    ge_study_query_gwas_lead <- paste0("https://www.ebi.ac.uk/eqtl/api/associations?variant_id=", variant_id,
                                       "&p_upper=", p_upper,
                                       "&study=", study_name,
                                       "&quant_method=", quant_method)
    fetched_df <- fetch_from_eqtl_cat_API(ge_study_query_gwas_lead)
    rnaseq_studies_sign_assocs <- rnaseq_studies_sign_assocs %>% rbind(fetched_df)
  }
  rnaseq_studies_sign_assocs$quant_method <- quant_method

  return(rnaseq_studies_sign_assocs)
}
```

Method to get all associations in cis region of each significant QTL
```{r}
get_assocs_in_region <- function(sign_assoc) {
  message(sign_assoc["qtl_group"])
  fetch_region_for_study_gene_link <- paste0("http://www.ebi.ac.uk/eqtl/api/chromosomes/",sign_assoc["chromosome"],
                                          "/associations?paginate=False&study=",sign_assoc["study_id"],
                                          "&qtl_group=",sign_assoc["qtl_group"],
                                          "&molecular_trait_id=",sign_assoc["molecular_trait_id"],
                                          "&quant_method=", sign_assoc["quant_method"],
                                          "&bp_lower=45980000&bp_upper=46200000&size=1000")
  
  message("Fetching: ", fetch_region_for_study_gene_link)
  region_for_study_gene_data <- fetch_from_eqtl_cat_API(fetch_region_for_study_gene_link)

  region_for_study_gene_data <- region_for_study_gene_data %>%
    dplyr::group_by(position) %>%
    dplyr::mutate(alt_allele_count = length(alt)) %>%
    dplyr::filter(alt_allele_count == 1)
  
  return(region_for_study_gene_data)
}
```

Method to perform colocalisation analysis.
```{r}
get_colocs <- function(eqtl_data_for_region, gwas_data_for_trait) {
  shared_positions = intersect(eqtl_data_for_region$position, gwas_data_for_trait$base_pair_location)

  eqtl_shared = dplyr::filter(eqtl_data_for_region, position %in% shared_positions) %>% 
    dplyr::mutate(variant_id = as.character(position))
  gwas_shared = dplyr::filter(gwas_data_for_trait, base_pair_location %in% shared_positions) %>% 
    dplyr::mutate(variant_id = as.character(base_pair_location))
  
  eQTL_dataset = list(pvalues = eqtl_shared$pvalue, 
                      N = (eqtl_shared$an)[1]/2, #The sample size of the eQTL dataset was 84
                      MAF = eqtl_shared$maf, 
                      type = "quant", 
                      beta = eqtl_shared$beta,
                      snp = eqtl_shared$variant_id)
  gwas_dataset = list(beta = gwas_shared$log_OR, #If log_OR column is full of NAs then use beta column instead
                      varbeta = gwas_shared$se^2, 
                      type = "cc", 
                      snp = gwas_shared$variant_id,
                      s = 0.5, #This is acutally not used, because we already specified varbeta above.
                      MAF = eqtl_shared$maf)
  
  coloc_res = coloc::coloc.abf(dataset1 = eQTL_dataset, dataset2 = gwas_dataset,p1 = 1e-4, p2 = 1e-4, p12 = 1e-5)
  return(coloc_res$summary)
}
```

Util method for saving plots in different formats
```{r}
save_ggplots <- function(plot, path = ".", filename = "unnamed_plot", height = 15, width = 15){
  ggsave(plot = plot,
       filename = paste0(filename, ".eps"), 
       path = path,
       device = "eps", 
       height = height, 
       width = width,
       units = "cm",
       dpi = 300)

ggsave(plot = plot,
       filename = paste0(filename, ".png"), 
       path = path,
       device = "png", 
       height = height, 
       width = width,
       units = "cm",
       dpi = 300)

ggsave(plot = plot,
       filename = paste0(filename, ".pdf"), 
       path = path,
       device = "pdf", 
       height = height, 
       width = width,
       units = "cm",
       dpi = 300)
}
```

Method to prepare data for plotting a faceted figure
```{r}
make_assocs_list_to_merged_plottable <- function(all_coloc_dt, quant_method = "ge"){
  merged_assoc_data <- data.frame()
  for (index in 1:(all_coloc_dt$eqtl_assocs_in_region %>% length())) {
    assoc_df <- all_coloc_dt$eqtl_assocs_in_region[[index]]
    assoc_df$coloc_PP4 <- round(all_coloc_dt$colocs[6,index], 3)
    assoc_df$coloc_PP3 <- round(all_coloc_dt$colocs[5,index], 3)
    assoc_df$track <- paste0("eQTL Catalogue\n", 
                             assoc_df$study_id, "_",quant_method,"\n", 
                             assoc_df$qtl_group,"\n",assoc_df$molecular_trait_id)
    
    if (index==1) {
      merged_assoc_data <- assoc_df
    } else {
      merged_assoc_data <- merged_assoc_data %>% rbind(assoc_df)  
    }
  }
  return(merged_assoc_data)
}
```


```{r}
plot_faceted_multi_manhattans <- function(merged_eqtl_assocs,
                                          gwas_data_for_trait,
                                          save_plot = FALSE,
                                          save_dir = ".",
                                          save_filename = "new_manhattan_plot",
                                          save_width = 15,
                                          save_height = NA,
                                          no_GWAS_plot = FALSE) {
  # get shared positions between GWAS and eQTL data
  shared_positions = intersect(merged_eqtl_assocs$position,
                               gwas_data_for_trait$base_pair_location)
  
  eqtl_shared = dplyr::filter(merged_eqtl_assocs, position %in% shared_positions) %>%
    dplyr::mutate(variant_id = as.character(position))
  gwas_shared = dplyr::filter(gwas_data_for_trait, base_pair_location %in% shared_positions) %>% 
    dplyr::mutate(variant_id = as.character(base_pair_location))
  
  merged_data_for_plot <- as.data.frame(eqtl_shared %>% dplyr::select(pvalue, position, track))
  if (!no_GWAS_plot) {
    gwas_shared$track <- "GWAS RA"
    gwas_sagred_trans <- gwas_shared %>% 
      dplyr::select(p_value, base_pair_location, track) %>% 
      stats::setNames(c("pvalue", "position", "track"))
    merged_data_for_plot <- merged_data_for_plot %>% rbind(gwas_sagred_trans)
    
    merged_data_for_plot$track <- factor(merged_data_for_plot$track) %>% 
    forcats::fct_relevel("GWAS RA", after = 0)
  }   
  
  region_coords = c(45980000, 46200000)
  plot_base_all = ggplot(merged_data_for_plot, aes_(x = ~ position, y = ~
                                                      -log(pvalue, 10))) + geom_blank()

  plot_gwas_eqtl_plot = plot_base_all +
    geom_point(color = "deepskyblue4") +
    theme_light() +
    ylab(expression(paste("-", log[10], " p-value"))) +
    scale_x_continuous(limits = region_coords, expand = c(0, 0)) +
    facet_grid(track ~ ., scales = "free_y") +
    theme(
      plot.margin = unit(c(0.1, 1, 0.1, 1), "line"),
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "none",
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text.y = element_text(colour = "grey10"),
      strip.background = element_rect(fill = "grey85")
    )
  
  summary_coloc <- merged_eqtl_assocs %>% 
    group_by(track) %>% 
    summarise(coloc_PP4 = unique(coloc_PP4), coloc_PP3 = unique(coloc_PP3))
  
  dat_text <- data.frame(
    label = c(paste0("PP4: ", summary_coloc$coloc_PP4, 
                   "\nPP3: ", summary_coloc$coloc_PP3)),
    track = c( summary_coloc$track))

  if (!no_GWAS_plot) {
    dat_text <- rbind(data.frame(label="", track="GWAS RA"), dat_text)
  }
  
  plot_gwas_eqtl <- plot_gwas_eqtl_plot + geom_text(
    data    = dat_text,
    mapping = aes(x = -Inf, y = -Inf, label = label),
    hjust   = -0.1,
    vjust = -3.4
  )
  
  if (save_plot) {
    save_height= ifelse(test = is.na(save_height), 
                        yes = ((as.integer(!no_GWAS_plot) + unique(merged_eqtl_assocs$track) %>% length()) * 4),
                        no = save_height)
    message("Saving: ", save_filename)
    save_ggplots(
      plot = plot_gwas_eqtl,
      path = save_dir,
      filename = save_filename,
      height = save_height,
      width = save_width
    )
  }
  return(plot_gwas_eqtl)
}
```

# Here we start the analysis

### We first fetch GWAS data
Chromosome: 20
study_accession: GCST002318
bp_lower: 45980000
bp_upper: 46200000
```{r fetch_GWAS_data}
RA_gwas_query_str <- "https://www.ebi.ac.uk/gwas/summary-statistics/api/chromosomes/20/associations?study_accession=GCST002318&bp_lower=45980000&bp_upper=46200000&size=1000"
gwas_data <- fetch_from_eqtl_cat_API(link = RA_gwas_query_str, is_gwas = TRUE)
message("Downloaded ", gwas_data %>% nrow(), " associations from GWAS Catalogue")
gwas_data <- gwas_data %>% 
  dplyr::filter(!ci_lower %>% is.na()) %>% 
  dplyr::filter(!ci_upper %>% is.na()) %>% 
  dplyr::mutate(log_OR = log(odds_ratio)) %>%
  dplyr::mutate(se = (log(ci_upper)-log(ci_lower))/3.92)
message("There remain ", gwas_data %>% nrow(), " associations after filtering invalid values")
```

Pick the lead variant from GWAS data
```{r}
lead_var_gwas = dplyr::arrange(gwas_data, p_value)[1,]
```

Pick significant associations in eQTL Catalogue with lead GWAS variant ID for RNA-seq and Microarray studies and merge two tables
```{r}
# "ROSMAP"
# "Schmiedel_2018",
rnaseq_study_names <- c("Alasoo_2018", "BLUEPRINT", "BrainSeq", "GENCORD", "GEUVADIS", 
                        "HipSci", "Lepik_2017", "Nedelec_2016", "Quach_2016", 
                        "Schwartzentruber_2018", "TwinsUK", "van_de_Bunt_2015")
rnaseq_studies_sign_assocs <- get_significant_assocs_of_var(study_ids = rnaseq_study_names, 
                                                            variant_id = lead_var_gwas$variant_id) 

microarray_study_names <- c("CEDAR", "Fairfax_2014", "Kasela_2017", "Naranbhai_2015", "Fairfax_2012")
microarray_studies_sign_assocs <- get_significant_assocs_of_var(study_ids = microarray_study_names, 
                                                                variant_id = lead_var_gwas$variant_id,
                                                                quant_method = "microarray")

all_studies_sign_assocs <- rnaseq_studies_sign_assocs %>% rbind(microarray_studies_sign_assocs)
```

## Plot pvalues of each significant association
```{r sign_QTLS_scatter_plot, message=FALSE, fig.height=7, fig.width=10}
# fetch all significant transcript QTLs
sign_tx_rs4239702 <-  get_significant_assocs_of_var(study_ids = rnaseq_study_names,
                                                    variant_id = lead_var_gwas$variant_id, 
                                                    quant_method = "tx")

all_studies_sign_assocs_with_tx <- all_studies_sign_assocs %>% rbind(sign_tx_rs4239702)
sign_assocs_to_plot_scatter <- all_studies_sign_assocs_with_tx %>% group_by(study_id, qtl_group) %>% 
  dplyr::filter(pvalue == min(pvalue)) %>% 
  summarise(condition_label = condition_label,  
            molecular_trait_id=molecular_trait_id, 
            pvalue = pvalue, 
            tissue_label = tissue_label, 
            quant_method = quant_method) %>% 
  arrange(pvalue) %>% 
  mutate(context = paste0(study_id,"_",condition_label))

sign_assocs_to_plot_scatter$quant_method[sign_assocs_to_plot_scatter$quant_method=="tx"] <- "RNA-seq Transcript usage"
sign_assocs_to_plot_scatter$quant_method[sign_assocs_to_plot_scatter$quant_method=="ge"] <- "RNA-seq Gene expression"
sign_assocs_to_plot_scatter$quant_method[sign_assocs_to_plot_scatter$quant_method=="microarray"] <- "Microarray Gene expression"
sign_assocs_to_plot_scatter$context <- factor(sign_assocs_to_plot_scatter$context, levels = rev(unique(sign_assocs_to_plot_scatter$context)))

sign_assocs_to_plot_scatter$tissue_label <- factor(sign_assocs_to_plot_scatter$tissue_label, levels = unique(sign_assocs_to_plot_scatter$tissue_label))

base_plot <- ggplot(sign_assocs_to_plot_scatter, aes(x=-log10(pvalue), y = context, color = tissue_label, shape = quant_method))
final_plot <- base_plot + geom_point() + theme_bw()+ scale_color_brewer(palette="Dark2") +
  xlab(expression(paste("-",log[10], " p-value"))) +
  ylab("Study and Condition") +
  labs(color = "Cell Type", shape = "Quantification method")

save_ggplots(plot = final_plot, path = "eQTL_API_usecase_figures", filename = "sign_eqtl_scatter_shape", height = 10, width = 18)
final_plot
```

Get all associations in cis region of each significant QTL
```{r get_associations_in_cis, message=FALSE}
all_coloc_data <- list()
all_coloc_data$eqtl_assocs_in_region <- apply(all_studies_sign_assocs, 1, get_assocs_in_region)
```

Perform colocalisation analysis.
```{r perform_coloc}
all_coloc_data$colocs <- sapply(all_coloc_data$eqtl_assocs_in_region, get_colocs, gwas_data_for_trait=gwas_data)
```

Prepare merged facet plottable dataframe
```{r merge_assoc_dataframes}
plottable_merged_data <- make_assocs_list_to_merged_plottable(all_coloc_data)
```

Plot faceted manhattan figure with multiple 
```{r plot_all_contexts_faceted, fig.width=8, fig.height=48}
all_plots_faceted <- plot_faceted_multi_manhattans(merged_eqtl_assocs = plottable_merged_data, 
                              gwas_data_for_trait = gwas_data,
                              save_plot = TRUE,
                              save_dir = "eQTL_API_usecase_figures", 
                              save_filename = "merged_manhattan", no_GWAS_plot = TRUE) 
all_plots_faceted
```

Plot specific figures in faceted plot
```{r plot_filtered_contexts_faceted, fig.width=8, fig.height=12}
plottable_merged_data_filt <- plottable_merged_data %>% 
  dplyr::filter(study_id %in% c("BLUEPRINT", "Quach_2016", "CEDAR", "Fairfax_2014")) %>% 
  dplyr::filter(qtl_group %in% c("monocyte_naive", "monocyte", "monocyte_CD14")) %>% 
  dplyr::filter(molecular_trait_id %in% c("ENSG00000101017", "ILMN_1779257"))

filt_plots_faceted <- plot_faceted_multi_manhattans(merged_eqtl_assocs = plottable_merged_data_filt, 
                              gwas_data_for_trait = gwas_data,
                              save_plot = TRUE,
                              save_dir = "eQTL_API_usecase_figures", 
                              save_filename = "merged_manhattan_filt", no_GWAS_plot = TRUE) 
filt_plots_faceted
```


## Analyse transcript usage associations in Alasoo_2018
```{r get_Alasoo_2018_transcript_QTLs, fig.width=8, fig.height=6}
sign_tx_Alasoo_2018 <- sign_tx_rs4239702 %>% dplyr::filter(study_id=="Alasoo_2018")

all_coloc_data_tx <- list()
all_coloc_data_tx$eqtl_assocs_in_region <- apply(sign_tx_Alasoo_2018, 1, get_assocs_in_region)

all_coloc_data_tx$colocs <- sapply(all_coloc_data_tx$eqtl_assocs_in_region, get_colocs, gwas_data_for_trait=gwas_data)
merged_plottable_df <-  make_assocs_list_to_merged_plottable(all_coloc_data_tx, quant_method = "tx")

manhattan_tx_gwas <- plot_faceted_multi_manhattans(merged_eqtl_assocs = merged_plottable_df, 
                                                 gwas_data_for_trait = gwas_data,
                                                 save_plot = TRUE,
                                                 save_dir = "eQTL_API_usecase_figures",
                                                 save_filename = "merged_manhattan_tx")
manhattan_tx_gwas
```
```{r add_CD40_transcript_location, fig.width=8, fig.height=3}
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl", version=78)

CD_40_ENST00000466205_exons <- getBM(attributes=c('ensembl_exon_id','exon_chrom_start','exon_chrom_end',
                                                  'strand', 'chromosome_name'), 
                                     filters = 'ensembl_transcript_id', 
                                     values ="ENST00000466205", 
                                     mart = ensembl)

CD_40_ENST00000466205_exons$strand <- "+"

exons_grange <- makeGRangesFromDataFrame(CD_40_ENST00000466205_exons, ignore.strand = FALSE, seqnames.field = "chromosome_name", start.field = 'exon_chrom_start', end.field = 'exon_chrom_end', strand.field = 'strand', keep.extra.columns = TRUE)
CD_40_exons_list <- GRangesList()
CD_40_exons_list[['ENST00000466205']] <- exons_grange

region_coords = c(45980000, 46200000)
CD40_exons_plot <- plotTranscripts(exons = CD_40_exons_list, rescale_introns = FALSE, region_coords = region_coords)
CD40_exons_plot
```

```{r plot_merged_manhattan_plot, fig.width=8, fig.height=18}
joint_plot = cowplot::plot_grid(manhattan_tx_gwas, filt_plots_faceted, CD40_exons_plot,
                                align = "v", ncol = 1, rel_heights = c(2,4, 1))

save_ggplots(plot = joint_plot, path = "eQTL_API_usecase_figures", filename = "new_tx_merged_plot_with_exon", height = 28, width = 15)
joint_plot
```



