---
title: "eQTL Catalogue coloc.susie example"
output: 
  html_document: 
    keep_md: yes
date: "2024-06-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Perform colocalisation with coloc.susie and lbf_variable files from the QTL Catalogue

Load required pacakges:
```{r}
library("coloc")
library("readr")
library("dplyr")
```

Extract BPI eQTL and pQTL LBF variables from eQTL Catalogue
```{bash}
#gunzip -c QTD000021.lbf_variable.txt.gz | head -n1 > QTD000021_ENSG00000101425.tsv && gunzip -c QTD000021.lbf_variable.txt.gz | grep ENSG00000101425 >> QTD000021_ENSG00000101425.tsv
#gunzip -c QTD000026.lbf_variable.txt.gz | head -n1 > QTD000026_ENSG00000101425.tsv && gunzip -c QTD000026.lbf_variable.txt.gz | grep ENSG00000101425 >> QTD000026_ENSG00000101425.tsv
#gunzip -c QTD000584.lbf_variable.txt.gz | head -n1 > QTD000584_BPI.tsv && gunzip -c QTD000584.lbf_variable.txt.gz | grep "BPI.4126.22.1..1" >> QTD000584_BPI.tsv
```

Import lbf variable values into R
```{r}
#BLUEPRINT monocytes
mono_lbf = readr::read_tsv("input_data/QTD000026_ENSG00000101425.tsv.gz", show_col_types = FALSE)

#BLUEPRINT neutrophils
neutrophil_lbf = readr::read_tsv("input_data/QTD000021_ENSG00000101425.tsv.gz", show_col_types = FALSE)

#INTERVAL plasma pQTLs
interval_lbf = readr::read_tsv("input_data/QTD000584_BPI.tsv.gz", show_col_types = FALSE)
```

Convert LBF data frames into matrixes suitable for the coloc.bf_bf() method

```{r}
#BLUEPRINT monocytes
mono_mat = as.matrix(dplyr::select(mono_lbf, lbf_variable1:lbf_variable10))
row.names(mono_mat) = mono_lbf$variant
mono_mat = t(mono_mat)

#BLUEPRINT neutrophils
neutro_mat = as.matrix(dplyr::select(neutrophil_lbf, lbf_variable1:lbf_variable10))
row.names(neutro_mat) = neutrophil_lbf$variant
neutro_mat = t(neutro_mat)

#INTERVAL plasma pQTLs
interval_mat = as.matrix(dplyr::select(interval_lbf, lbf_variable1:lbf_variable10))
row.names(interval_mat) = interval_lbf$variant
interval_mat = t(interval_mat)
```

### Perform all pairwise colocalisations

BLUEPRINT neutrophils eQTL vs INTERVAL plasma pQTL
```{r}
neutro_coloc = coloc::coloc.bf_bf(neutro_mat, interval_mat)
dplyr::as_tibble(neutro_coloc$summary) %>% dplyr::filter(PP.H4.abf > 0.8)
```


BLUEPRINT monocytes eQTL vs INTERVAL plasma pQTL
```{r}
mono_coloc = coloc::coloc.bf_bf(mono_mat, interval_mat)
dplyr::as_tibble(mono_coloc$summary) %>% dplyr::filter(PP.H4.abf > 0.8)
```


BLUEPRINT monocytes eQTL vs BLUEPRINT neutrophils eQTL
```{r}
eQTL_coloc = coloc::coloc.bf_bf(neutro_mat, mono_mat)
dplyr::as_tibble(eQTL_coloc$summary) %>% dplyr::filter(PP.H4.abf > 0.8)
```

